### 2018-06-11 paper/3-abstractions.tex
Finish S3

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
dependencies :: Task Applicative k v -> k -> [k]
dependencies task key = case task (\k -> Const [k]) key of
                            Nothing         -> []
                            Just (Const ks) -> ks
\end{minted}
\vspace{1mm}

Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-types}. We
instantiate \hs{f} to \hs{Const}~\hs{[@@k]}.  So a value of type \hs{f}~\hs{v},
or in this case \hs{Const}~\hs{[@@k]}~\hs{v}, contains no value \hs{v}, but does
contain a list of keys of type \hs{[@@k]} which we use to record dependencies.
The \hs{fetch} callback that we pass to \hs{task} records a single dependency;
and the standard definition of \hs{Applicative} for \hs{Const} (which we give
in Fig.~\ref{fig-types}) combines the dependencies from different parts of the
task. Running the task with \hs{f}~=~\hs{Const}~\hs{[@@k]} will thus
accumulate a list of the task's dependencies -- and that is just what
\hs{dependencies} does:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies sprsh1 "A1"
[@@]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies sprsh1 "B1"
["A1", "A2"]
\end{minted}
\vspace{1mm}

@@

So much for applicative tasks. What about monadic tasks with dynamic
dependencies? As we have seen in~\S\ref{sec-background-shake}, dynamic
dependencies need to be tracked too. This cannot be done statically; notice that
the application of the function \hs{dependencies} to a \hs{Task}~\hs{Monad} will
not typecheck. We need to run a monadic task on a store with concrete values,
which will determine the discovered dependencies. Accordingly, we introduce
the function \hs{track}: a combination of \hs{compute} and \hs{dependencies}
that computes both the resulting value and the list of its dependencies in an
arbitrary monadic context~\hs{m}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
import Control.Monad.Writer
\end{minted}
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
track :: Monad m => Task Monad k v -> (k -> m v) -> k -> Maybe (m (v, [k]))
track task fetch = fmap runWriterT . task trackingFetch
  where
    trackingFetch :: k -> WriterT [k] m v
    trackingFetch k = tell [k] >> lift (fetch k)
\end{minted}
\vspace{1mm}

\noindent
This implementation uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad}, which allows us
to record additional information -- a list of keys of type \hs{[@@k]} -- when
computing a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch}
with a \hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key. The \hs{task} returns the value of type
\hs{Maybe}~\hs{(@@WriterT}~\hs{[@@k]}~\hs{m}~\hs{v)}, which we unwrap by applying
\hs{runWriterT} to the contents of \hs{Maybe}.
Below we give an example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ fetchIO k = do putStr (k ++ ": "); read <$> getLine
@\ghci@ fromJust $ track sprsh2 fetchIO "B1"
C1: 1
B2: 10
(10,["C1","B2"])
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ fromJust $ track sprsh2 fetchIO "B1"
C1: 2
A2: 20
(20,["C1","A2"])
\end{minted}
\vspace{1mm}

>>>>

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
dependencies :: Task Applicative k v -> [k]
dependencies task = getConst $ run task (\k -> Const [k])
\end{minted}
\vspace{1mm}

\noindent
Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-types}. We
instantiate \hs{f} to \hs{Const}~\hs{[@@k]}. So a value of type \hs{f}~\hs{v},
or in this case \hs{Const}~\hs{[@@k]}~\hs{v}, contains no value \hs{v}, but does
contain a list of keys of type \hs{[@@k]} which we use to record dependencies.
The \hs{fetch} callback that we pass to \hs{task} records a single dependency;
and the standard definition of \hs{Applicative} for \hs{Const} (which we give
in Fig.~\ref{fig-types}) combines the dependencies from different parts of the
task. Running the task with \hs{f}~=~\hs{Const}~\hs{[@@k]} will thus
accumulate a list of the task's dependencies -- and that is just what
\hs{dependencies} does:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "A1"
[@@]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "B1"
["A1", "A2"]
\end{minted}
\vspace{1mm}

@@

So much for applicative tasks. What about monadic tasks with dynamic
dependencies? As we have seen in~\S\ref{sec-background-shake}, dynamic
dependencies need to be tracked too. This cannot be done statically; notice that
we cannot apply the function \hs{dependencies} to a \hs{Task}~\hs{Monad} because
the \hs{Const} functor has no \hs{Monad} instance. We need to run a monadic task
on a store with concrete values, which will determine the discovered
dependencies. Accordingly, we introduce the function \hs{track}: a combination
of \hs{compute} and \hs{dependencies} that computes both the resulting value and
the list of its dependencies in an arbitrary monadic context~\hs{m}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
import Control.Monad.Writer
\end{minted}
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
track :: Monad m => Task Monad k v -> (k -> m v) -> m (v, [k])
track task fetch = runWriterT $ run task trackingFetch
  where
    trackingFetch :: k -> WriterT [k] m v
    trackingFetch k = tell [k] >> lift (fetch k)
\end{minted}
\vspace{1mm}

\noindent
This implementation uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad}, which allows us
to record additional information -- a list of keys of type \hs{[@@k]} -- when
computing a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch}
with a \hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key. The \hs{task} returns the value of type
\hs{WriterT}~\hs{[@@k]}~\hs{m}~\hs{v}, which we unwrap by \hs{runWriterT}. Here
is an example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ fetchIO k = do putStr (k ++ ": "); read <$> getLine
@\ghci@ track (fromJust $ sprsh2 "B1") fetchIO
C1: 1
B2: 10
(10,["C1","B2"])
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ track (fromJust $ sprsh2 "B1") fetchIO
C1: 2
A2: 20
(20,["C1","A2"])
\end{minted}
\vspace{1mm}
